---
title: "Psychometric Network Analysis of Taste Schemas"
author: "Omar Lizardo"
date: today
format: 
  html:
    toc: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Setup

```{r}
#| label: setup
library(EGAnet)
library(corclass)
library(ggridges)
library(haven)
library(here)
library(Hmisc)
library(patchwork)
library(qgraph)
library(tidyverse)
library(mice)
library(networktools)
```

## Data Import and Cleaning

We load the raw data, isolate the taste variables, and apply our reverse-coding scheme so that 1 = Dislike, 2 = Mixed, 3 = Like. We also drop respondents who have zero variance across their taste evaluations.

```{r}
#| label: data-cleaning
set.seed(1234)

# Define column mapping for clarity
taste_cols <- c(
  "Classical" = "classicaltaste", "Opera" = "operataste", "Jazz" = "jazztaste", 
  "Musicals" = "bwaysttaste", "Easy" = "moodeztaste", "Big_Band" = "bbandtaste",
  "Classic_Rock" = "croldtaste", "Country" = "countrytaste", "Bluegrass" = "bluegtaste", 
  "Folk" = "folktaste", "Gospel" = "hymgostaste", "Latin" = "latspsaltaste",
  "Rap" = "raphiphoptaste", "Blues_RandB" = "blurbtaste", 
  "Reggae" = "reggaetaste", "Pop" = "toppoptaste", "Cont_Rock" = "controcktaste", 
  "Alt_Rock" = "indalttaste", "Dance" = "danclubtaste", "Heavy_Metal" = "hvymtltaste"
)

raw_data <- read_dta(here("data", "SSI2012.dta"))

taste_dat <- raw_data |> 
  select(id, all_of(taste_cols)) |> 
  # Recode so 1=Dislike, 2=Mixed, 3=Like. 4 means missing.
  mutate(across(-id, ~ case_when(
    as.integer(.x) == 1L ~ 3L,
    as.integer(.x) == 2L ~ 1L,
    as.integer(.x) == 3L ~ 2L,
    TRUE ~ NA_integer_
  ))) |> 
  # Calculate variance/SD to drop zero-variance rows
  mutate(taste_sd = apply(pick(-id), 1, sd, na.rm = TRUE)) |> 
  filter(!is.na(taste_sd) & taste_sd != 0) |> 
  select(-taste_sd)
```

## Imputation

Instead of taking the median across 10 imputed datasets (which breaks Rubin's rules and can create non-integer ordinals), we use `mice` to generate a single robust imputation (using Predictive Mean Matching) solely to create a complete dataset for the correlational class analysis.

```{r}
#| label: imputation
imp_res <- mice(taste_dat |> select(-id), m = 1, method = "pmm", seed = 1234, printFlag = FALSE)
taste_dat_imp <- complete(imp_res, 1)

# Attach ID back
taste_dat_imp <- bind_cols(id = taste_dat$id, taste_dat_imp)
```

## Correlational Class Analysis (CCA)

```{r}
#| label: cca-analysis
cca_res <- cca(taste_dat_imp |> select(-id), filter.value = 0.01)

taste_dat_c <- taste_dat_imp |> 
  mutate(Class = as.factor(cca_res$membership))

# Quick count of classes
table(taste_dat_c$Class)
```

## Network Estimation & Centrality

We iterate over the classes to estimate the networks dynamically. `EGA` will default to polychoric correlations for these ordinal variables.

```{r}
#| label: ega-and-centrality
# Split data by class
class_data_list <- taste_dat_c |> 
  group_split(Class) |> 
  set_names(paste0("Class_", sort(unique(taste_dat_c$Class))))

# Function to estimate EGA and extract centrality
estimate_ega <- function(dat) {
  net_dat <- dat |> select(-id, -Class)
  # Use Leiden algorithm for community detection
  ega_obj <- EGA(net_dat, algorithm = "leiden", objective_function = "CPM", plot.EGA = FALSE)
  
  list(ega = ega_obj, data = net_dat)
}

# Apply to all classes
results_classes <- map(class_data_list, estimate_ega)
```

### Network Plots

We can visualize the underlying graph structure for each schema. The nodes (genres) are grouped and colored based on the communities identified by the Leiden algorithm.

```{r}
#| label: network-plots
#| fig-width: 16
#| fig-height: 10
plot_network <- function(res, title) {
  # The plot method for EGA objects returns a ggplot object
  plot(res$ega, plot.args = list(vsize = 6, label.size = 4, alpha = 0.8)) +
    labs(title = title) +
    theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
          legend.position = "bottom")
}

net_plots <- imap(results_classes, ~ plot_network(.x, .y))
wrap_plots(net_plots, ncol = 3)
```

### Centrality Plots

By standardizing centrality measures to Z-scores, we can directly compare Strength, Closeness, and Betweenness on the same scale. The genres are ordered by Strength, identifying the most core components of each schema.

```{r}
#| label: centrality-plots
#| fig-width: 16
#| fig-height: 10
plot_centrality <- function(res, title) {
  # qgraph::centralityPlot calculates Z-scores and returns a ggplot object
  centralityPlot(res$ega$network, 
                 include = c("Strength", "Closeness", "Betweenness"),
                 scale = "z-scores",
                 orderBy = "Strength") +
    labs(title = title) +
    theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
}

cent_plots <- imap(results_classes, ~ plot_centrality(.x, .y))
wrap_plots(cent_plots, ncol = 3)
```

### Schema Interpretations

Based on the psychometric network structures and node centralities, we can interpret these schematic classes not just by *what* respondents like or dislike, but by how their tastes are *organized*. Genres with high **Strength** act as core anchors—meaning feelings about that genre strongly dictate feelings about the rest of the music landscape. **Betweenness** highlights genres that act as bridges connecting disparate clusters.

* **Schema 1: The "Highbrow / Rock Anchor" Schema**
  Core Anchors: `Classical`, `Cont_Rock`, `Classic_Rock`, `Jazz`. For this group, "highbrow" genres and standard rock act as the central organizing principles. Their opinions on these genres strongly predict their entire taste profile. They tend to strongly reject Rap and Opera.
* **Schema 2: The "Classic Rock / Pop Core" Schema**
  Core Anchors: `Classic_Rock`, `Rap`, `Bluegrass`. This schema represents a very common taste profile anchored centrally by Classic Rock. Interestingly, `Rap` has very high strength and betweenness here—meaning that whether they tolerate or reject Rap is a massive polarizing divider that structurally dictates the rest of their network.
* **Schema 3: The "Pop / Contemporary Bridge" Schema**
  Core Anchors: `Pop`, `Cont_Rock`, `Opera`, `Rap`. Organized heavily around mainstream pop and contemporary music, `Pop` is the absolute central load-bearing pillar. `Opera` acts as a strong negative anchor (highly central because almost everyone in this class uses their uniform dislike of it to structurally define their taste).
* **Schema 4: The "Easy Listening / Roots" Schema**
  Core Anchors: `Easy` (Easy Listening), `Pop`, `Bluegrass`, `Jazz`. A softer, traditionalist schema. Harder genres like Heavy Metal are pushed completely to the network's periphery. The network revolves structurally around accessible, easy-to-digest music and roots music.
* **Schema 5: The "Polarized / Omnivore" Schema**
  Core Anchors: `Rap`, `Big_Band`, `Cont_Rock`, `Folk`. The most eclectic and structurally diverse schema. The core anchors span across completely different historical and cultural eras. High betweenness for these genres means this class evaluates music across boundaries, heavily indicating an "Omnivore" profile where genres serve as bridges rather than isolated islands.

***

### Genre Evaluations (Ridge Plots)

```{r}
#| label: ridge-plots
#| fig-width: 16
#| fig-height: 4
taste_long <- taste_dat_c |> 
  pivot_longer(cols = -c(id, Class), names_to = "genre", values_to = "eval")

plot_ridges <- function(df, title) {
  ggplot(df, aes(x = eval, y = reorder(genre, eval))) + 
    geom_density_ridges2() + 
    scale_x_continuous(breaks = 1:3, labels = c('Dislike','Mixed','Like')) + 
    labs(x = "", y = "", title = title) +
    theme_minimal()
}

ridge_full <- plot_ridges(taste_long, "Full Sample")
ridge_classes <- map(sort(unique(taste_dat_c$Class)), ~ {
  plot_ridges(taste_long |> filter(Class == .x), paste("Schema", .x))
})

wrap_plots(c(list(ridge_full), ridge_classes), nrow = 1)
```

## Validation Step

To structurally validate our uncovered schemas, we assess whether the psychological organization of tastes (network centrality) corresponds to behavioral listening patterns. We correlate the **Expected Influence** (node strength) of each genre within its respective schema network to the genre's correlation with the respondent's overall **"omnivorousness"** (the total count of genres they actually listen to).

If the schema networks are capturing genuine cultural logic, genres that anchor the schema (high expected influence) should strongly predict an individual's behavioral omnivorousness.

```{r}
#| label: validation-step
#| fig-width: 14
#| fig-height: 6

# Load behavioral listening data 
beh_cols <- c(
  "Classical" = "classicallis", "Opera" = "operalis", "Jazz" = "jazzlis", 
  "Musicals" = "bwaystlis", "Easy" = "moodezlis", "Big_Band" = "bbandlis",
  "Classic_Rock" = "croldlis", "Country" = "countrylis", "Bluegrass" = "blueglis", 
  "Folk" = "folklis", "Gospel" = "hymgoslis", "Latin" = "latspsallis",
  "Rap" = "raphiphoplis", "Blues_RandB" = "blurblis", 
  "Reggae" = "reggaelis", "Pop" = "toppoplis", "Cont_Rock" = "controcklis", 
  "Alt_Rock" = "indaltlis", "Dance" = "danclublis", "Heavy_Metal" = "hvymtllis"
)

lis_dat <- raw_data |> 
  select(id, all_of(beh_cols)) |> 
  drop_na() |> 
  mutate(omni = rowSums(across(-id)))

# Calculate validation metrics per class
calc_validation <- function(class_idx, list_obj) {
  ei_res <- expectedInf(list_obj$ega$network)$step1
  ei_df <- tibble(genre = names(ei_res), exp_inf = scale(ei_res)[,1])
  
  dat <- taste_dat_c |> 
    filter(Class == class_idx) |> 
    inner_join(lis_dat |> select(id, omni), by = "id")
  
  corrs <- map_dbl(names(ei_res), ~ cor.test(dat[[.x]], dat$omni, method = "spearman")$estimate)
  
  ei_df |> mutate(omni_corr = scale(corrs)[,1], Class = paste("Schema", class_idx))
}

val_df <- map2_dfr(sort(unique(taste_dat_c$Class)), results_classes, calc_validation)

ggplot(val_df, aes(x = exp_inf, y = omni_corr, color = Class)) +
  geom_smooth(method = "lm", alpha = 0.15) +
  ggrepel::geom_text_repel(aes(label = genre), size = 3) +
  facet_wrap(~ Class, nrow = 1) +
  theme_classic() + 
  geom_hline(yintercept = 0, color = "gray") + 
  geom_vline(xintercept = 0, color = "gray") +
  labs(x = "Expected Influence (Z)", y = "Correlation with Omnivorousness (Z)") +
  theme(legend.position = "none")
```
